---
title: "Check $F_{is}$ and LD of microsatellite loci"
author: "Jenna Melanson"
date: "2025-06-20"
output:
  pdf_document: default
  html_document: default
bibliography: "/Users/jenna1/Documents/UBC/bombus_project/fv_landscapeforaging/docs/phd_bombus.bib"
---

```{r setup, include=FALSE, cache = TRUE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/Users/jenna1/Documents/UBC/bombus_project/fv_landscapeforaging")

# prep workspace
rm(list = ls())
setwd("/Users/jenna1/Documents/UBC/bombus_project/fv_landscapeforaging")

# first, load in packages
source('colony_assignments/src/init.R')
source('colony_assignments/src/colony_assignment_functions.R')
library(dplyr)
library(tidyr)
library(stringr)
library(rcolony)
library(genepop)
library(data.table)
library(cowplot)
library(raster)
library(lmerTest)
library(lme4)
library(emmeans)
library(ggplot2)
library(gridExtra)
library(igraph)
library(adegenet)
library(devtools)
library(PopGenUtils)
library(polysat)
library(hierfstat)


# load in raw data

# for mixtus
mtestalleles2022 = read.csv("data/merged_by_year/csvs/mixtus_2022_scores.csv")
mtestalleles2022 = mtestalleles2022[,-1]
mtestalleles2023 = read.csv("data/merged_by_year/csvs/mixtus_2023_scores.csv")
mtestalleles2023 = mtestalleles2023[,-1]

# for impatiens
itestalleles2022 = read.csv("data/merged_by_year/csvs/impatiens_2022_scores.csv")
itestalleles2022 = itestalleles2022[,-1]
itestalleles2023 = read.csv("data/merged_by_year/csvs/impatiens_2023_scores.csv")
itestalleles2023 = itestalleles2023[,-1]
```

## Remove siblings before marker assessment
We used an iterative approach to assign workers and queens to their natal colonies. First, using the pedigree reconstruction software COLONY 2.0 [@jonesCOLONYProgramParentage2010], we assigned full siblingships based on all available microsatellite data, assuming male and female monogamy and no inbreeding. A single run was carried out for each species and year, using the software's full-likelihood approach and no siblingship size scaling or priors. Siblingships were maintained at this stage only if P(full sibling dyad) = 1. A single individual from each putative colony (including non-circular colonies) was maintained for downstream analyses of microsatellite locus quality.

```{r remove sibs before genepop, include = FALSE, cache = TRUE}

#################################
# IMPATIENS 2022
################################

# load in FS dyad files
fsDyad2022 = trimws(readLines("colony_assignments/Colony2/initial/impatiens2022.FullSibDyad"))
split_dyads2022 <- strsplit(fsDyad2022, ",")
bestdyads2022 <- do.call(rbind, split_dyads2022)
colnames(bestdyads2022) <- bestdyads2022[1, ]
bestdyads2022 <- as.data.frame(bestdyads2022[-1, ])
bestdyads2022$Probability = as.numeric(bestdyads2022$Probability)

# remove sibpairs below threshold probability (p = 1)
bestdyads2022_filtered = bestdyads2022 %>% filter(Probability == 1)

# make an undirected graph of families
graph_22 <- graph_from_data_frame(bestdyads2022_filtered[, c("OffspringID1", "OffspringID2")], directed = FALSE)

# find connected components (families)
components <- components(graph_22)

# assign family ID to each individual
membership_df <- data.frame(
  OffspringID = names(components$membership),
  Family = components$membership
)

#maintain a single sib per putative colony
sibs_to_keep = membership_df %>%
  group_by(Family) %>%
  sample_n(1)
sibs_to_remove = membership_df$OffspringID[!membership_df$OffspringID %in% sibs_to_keep$OffspringID]
itestalleles2022 = itestalleles2022[!itestalleles2022$barcode_id %in% sibs_to_remove,]


#################################
# IMPATIENS 2023
################################

# load in FS dyad files
fsDyad2023 = trimws(readLines("colony_assignments/Colony2/initial/impatiens2023.FullSibDyad"))
split_dyads2023 <- strsplit(fsDyad2023, ",")
bestdyads2023 <- do.call(rbind, split_dyads2023)
colnames(bestdyads2023) <- bestdyads2023[1, ]
bestdyads2023 <- as.data.frame(bestdyads2023[-1, ])
bestdyads2023$Probability = as.numeric(bestdyads2023$Probability)

# remove sibpairs below threshold probability (p = 1)
bestdyads2023_filtered = bestdyads2023 %>% filter(Probability == 1)

# make an undirected graph of families
graph_22 <- graph_from_data_frame(bestdyads2023_filtered[, c("OffspringID1", "OffspringID2")], directed = FALSE)

# find connected components (families)
components <- components(graph_22)

# assign family ID to each individual
membership_df <- data.frame(
  OffspringID = names(components$membership),
  Family = components$membership
)

#maintain a single sib per putative colony
sibs_to_keep = membership_df %>%
  group_by(Family) %>%
  sample_n(1)
sibs_to_remove = membership_df$OffspringID[!membership_df$OffspringID %in% sibs_to_keep$OffspringID]
itestalleles2023 = itestalleles2023[!itestalleles2023$barcode_id %in% sibs_to_remove,]



#################################
# MIXTUS 2022
################################
# load in FS dyad files
fsDyad2022 = trimws(readLines("colony_assignments/Colony2/initial/mixtus2022.FullSibDyad"))
split_dyads2022 <- strsplit(fsDyad2022, ",")
bestdyads2022 <- do.call(rbind, split_dyads2022)
colnames(bestdyads2022) <- bestdyads2022[1, ]
bestdyads2022 <- as.data.frame(bestdyads2022[-1, ])
bestdyads2022$Probability = as.numeric(bestdyads2022$Probability)

# remove sibpairs below threshold probability (p = 1)
bestdyads2022_filtered = bestdyads2022 %>% filter(Probability == 1)

# make an undirected graph of families
graph_22 <- graph_from_data_frame(bestdyads2022_filtered[, c("OffspringID1", "OffspringID2")], directed = FALSE)

# find connected components (families)
components <- components(graph_22)

# assign family ID to each individual
membership_df <- data.frame(
  OffspringID = names(components$membership),
  Family = components$membership
)

#maintain a single sib per putative colony
sibs_to_keep = membership_df %>%
  group_by(Family) %>%
  sample_n(1)
sibs_to_remove = membership_df$OffspringID[!membership_df$OffspringID %in% sibs_to_keep$OffspringID]
mtestalleles2022 = mtestalleles2022[!mtestalleles2022$barcode_id %in% sibs_to_remove,]



#################################
# MIXTUS 2023
################################
# load in FS dyad files
fsDyad2023 = trimws(readLines("colony_assignments/Colony2/initial/mixtus2023.FullSibDyad"))
split_dyads2023 <- strsplit(fsDyad2023, ",")
bestdyads2023 <- do.call(rbind, split_dyads2023)
colnames(bestdyads2023) <- bestdyads2023[1, ]
bestdyads2023 <- as.data.frame(bestdyads2023[-1, ])
bestdyads2023$Probability = as.numeric(bestdyads2023$Probability)

# remove sibpairs below threshold probability (p = 1)
bestdyads2023_filtered = bestdyads2023 %>% filter(Probability == 1)

# make an undirected graph of families
graph_22 <- graph_from_data_frame(bestdyads2023_filtered[, c("OffspringID1", "OffspringID2")], directed = FALSE)

# find connected components (families)
components <- components(graph_22)

# assign family ID to each individual
membership_df <- data.frame(
  OffspringID = names(components$membership),
  Family = components$membership
)

#maintain a single sib per putative colony
sibs_to_keep = membership_df %>%
  group_by(Family) %>%
  sample_n(1)
sibs_to_remove = membership_df$OffspringID[!membership_df$OffspringID %in% sibs_to_keep$OffspringID]
mtestalleles2023 = mtestalleles2023[!mtestalleles2023$barcode_id %in% sibs_to_remove,]
```

## Check distribution of $F_{is}$ estimates across loci
While COLONY 2.0 can account for inbreeding at the population level, locus-specific estimates of $F_{is}$ should generally be similar to one another, reflecting their shared evolutionary history. Loci with $F_{is}$ estimates that deviate significantly from the species mean (across loci) may suffer from null alleles or other types of scoring errors that can bias siblingship assignment. We therefore tested individual locus deviations from population mean $F_{is}$ as a criterion for marker inclusion/exclusion.

Single locus $F_{is}$ estimates were computed for all site, year, species groups following @weirEstimatingFStatisticsAnalysis1984 in the package *genepop* [@roussetGenepop007CompleteReimplementation2008]. The distributions of $F_{is}$ estimates for each species at each site are shown below. The blue dotted line indicates $F_{is}$ = 0, and the red dotted lines indicate $F_{is}$ = $\pm$ 0.1.

```{r calculate Fis and LD, include = FALSE, cache = TRUE}
######################################
# Prep BOMBUS MIXTUS for genepop
#####################################

#combine columns for each locus
ms_m22 = data.table(mtestalleles2022)
ms_m22[ms_m22 ==0] <- "000"


i1 <- seq(1, length(ms_m22)-1, 2)
i2 <- seq(2, length(ms_m22)-1, 2)
msgenepop_m22 = ms_m22[, Map(paste,
                             .SD[, i1, with = FALSE], .SD[, i2, with = FALSE], 
                             MoreArgs = list(sep="")), 
                       by = "barcode_id"]
colnames(msgenepop_m22) = c("barcode_id", "BT10", "BTMS0104", "BTMS0057", "BTMS0086",
                            "BTMS0066", "BTMS0062", "BTMS0136","BTERN01", "BTMS0126", 
                            "BTMS0059", "BL13", "BTMS0083", "B126")

#sort allele table by barcode ID
msgenepop_m22 <- msgenepop_m22[order(msgenepop_m22$barcode_id),]
msgenepop_m22$site = substring(msgenepop_m22$barcode_id, 1, 1)

haplotypes_m22 <- as.data.frame(paste(msgenepop_m22$site, ","," ", 
                                      msgenepop_m22$BT10," ", msgenepop_m22$BTMS0104, " ",
                                      msgenepop_m22$BTMS0057," ", msgenepop_m22$BTMS0086, " ",
                                      msgenepop_m22$BTMS0066," ", msgenepop_m22$BTMS0062, " ",
                                      msgenepop_m22$BTMS0136," ", msgenepop_m22$BTERN01, " ",
                                      msgenepop_m22$BTMS0126," ", msgenepop_m22$BTMS0059," ",
                                      msgenepop_m22$BL13," ", msgenepop_m22$BTMS0083, " ",
                                      msgenepop_m22$B126,
                                      sep = ""))

#Build the Genepop format
sink("data/genepop/ms_genepop_format_mixtus2022.txt")
cat("Title: B. mixtus workers (2022) \n")
cat("BT10 \n")
cat("BTMS0104 \n")
cat("BTMS0057 \n")
cat("BTMS0086 \n")
cat("BTMS0066 \n")
cat("BTMS0062 \n")
cat("BTMS0136 \n")
cat("BTERN01 \n")
cat("BTMS0126 \n")
cat("BTMS0059 \n")
cat("BL13 \n")
cat("BTMS0083 \n")
cat("B126 \n")
cat("Pop \n")
invisible(apply(as.data.frame(haplotypes_m22[substring(haplotypes_m22[,1],1,1) == "E",]), 1,function(x) cat(x,"\n")))
cat("Pop \n")
invisible(apply(as.data.frame(haplotypes_m22[substring(haplotypes_m22[,1],1,1) == "W",]), 1,function(x) cat(x,"\n")))
cat("Pop \n")
invisible(apply(as.data.frame(haplotypes_m22[substring(haplotypes_m22[,1],1,1) == "S",]), 1,function(x) cat(x,"\n")))
cat("Pop \n")
invisible(apply(as.data.frame(haplotypes_m22[substring(haplotypes_m22[,1],1,1) == "N",]), 1,function(x) cat(x,"\n")))
cat("Pop \n")
invisible(apply(as.data.frame(haplotypes_m22[substring(haplotypes_m22[,1],1,1) == "P",]), 1,function(x) cat(x,"\n")))
cat("Pop \n")
invisible(apply(as.data.frame(haplotypes_m22[substring(haplotypes_m22[,1],1,1) == "H",]), 1,function(x) cat(x,"\n")))
cat("Pop \n")
sink()


######################################
# Run genepop on samples
#####################################

#test for HWE
test_HW("data/genepop/ms_genepop_format_mixtus2022.txt", which="Proba", "data/genepop/mixtusHWE2022.txt.D")

#test for LD
test_LD("data/genepop/ms_genepop_format_mixtus2022.txt","data/genepop/mixtusLD2022.txt.DIS")


######################################
# Repeat for 2023 mixtus !
#####################################

##### Prepare alleles table for genepop
#combine columns for each locus
ms_m23 = data.table(mtestalleles2023)
ms_m23[ms_m23 ==0] <- "000"

i1 <- seq(1, length(ms_m23)-1, 2)
i2 <- seq(2, length(ms_m23)-1, 2)
msgenepop_m23 = ms_m23[, Map(paste,
                             .SD[, i1, with = FALSE], .SD[, i2, with = FALSE], 
                             MoreArgs = list(sep="")), 
                       by = "barcode_id"]
colnames(msgenepop_m23) = c("barcode_id", "BT10", "BTMS0104", "BTMS0057", "BTMS0086",
                            "BTMS0066", "BTMS0062", "BTMS0136","BTERN01", "BTMS0126", 
                            "BTMS0059", "BL13", "BTMS0083", "B126")


#select and join alleles

#sort allele table by barcode ID
msgenepop_m23 <- msgenepop_m23[order(msgenepop_m23$barcode_id),]
msgenepop_m23$site = substring(msgenepop_m23$barcode_id, 1, 1)

haplotypes_m23 <- as.data.frame(paste(msgenepop_m23$site, ","," ", 
                                      msgenepop_m23$BT10," ", msgenepop_m23$BTMS0104, " ",
                                      msgenepop_m23$BTMS0057," ", msgenepop_m23$BTMS0086, " ",
                                      msgenepop_m23$BTMS0066," ", msgenepop_m23$BTMS0062, " ",
                                      msgenepop_m23$BTMS0136," ", msgenepop_m23$BTERN01, " ",
                                      msgenepop_m23$BTMS0126," ", msgenepop_m23$BTMS0059," ",
                                      msgenepop_m23$BL13," ", msgenepop_m23$BTMS0083, " ",
                                      msgenepop_m23$B126,
                                      sep = ""))

#Build the Genepop format
sink("data/genepop/ms_genepop_format_mixtus2023.txt")
cat("Title: B. mixtus workers (2023) \n")
cat("BT10 \n")
cat("BTMS0104 \n")
cat("BTMS0057 \n")
cat("BTMS0086 \n")
cat("BTMS0066 \n")
cat("BTMS0062 \n")
cat("BTMS0136 \n")
cat("BTERN01 \n")
cat("BTMS0126 \n")
cat("BTMS0059 \n")
cat("BL13 \n")
cat("BTMS0083 \n")
cat("B126 \n")
cat("Pop \n")
invisible(apply(as.data.frame(haplotypes_m23[substring(haplotypes_m23[,1],1,1) == "E",]), 1,function(x) cat(x,"\n")))
cat("Pop \n")
invisible(apply(as.data.frame(haplotypes_m23[substring(haplotypes_m23[,1],1,1) == "W",]), 1,function(x) cat(x,"\n")))
cat("Pop \n")
invisible(apply(as.data.frame(haplotypes_m23[substring(haplotypes_m23[,1],1,1) == "S",]), 1,function(x) cat(x,"\n")))
cat("Pop \n")
invisible(apply(as.data.frame(haplotypes_m23[substring(haplotypes_m23[,1],1,1) == "N",]), 1,function(x) cat(x,"\n")))
cat("Pop \n")
invisible(apply(as.data.frame(haplotypes_m23[substring(haplotypes_m23[,1],1,1) == "P",]), 1,function(x) cat(x,"\n")))
cat("Pop \n")
invisible(apply(as.data.frame(haplotypes_m23[substring(haplotypes_m23[,1],1,1) == "H",]), 1,function(x) cat(x,"\n")))
cat("Pop \n")
sink()


######################################
# Run genepop on samples
#####################################

#test for HWE
test_HW("data/genepop/ms_genepop_format_mixtus2023.txt", 
        which="Proba", 
        "data/genepop/mixtusHWE2023.txt.D")

#test for LD
test_LD("data/genepop/ms_genepop_format_mixtus2023.txt",
        "data/genepop/mixtusLD2023.txt.DIS")


######################################
# Repeat for BOMBUS IMPATIENS
######################################

######################################
# Prep alleles table for genepop
#####################################

### first for 2022! 
#combine columns for each locus
ms_i22 = data.table(itestalleles2022)
ms_i22[ms_i22 ==0] <- "000"


i1 <- seq(1, length(ms_i22)-1, 2)
i2 <- seq(2, length(ms_i22)-1, 2)
msgenepop_i22 = ms_i22[, Map(paste,
                         .SD[, i1, with = FALSE], .SD[, i2, with = FALSE], 
                         MoreArgs = list(sep="")), 
                   by = "barcode_id"]
colnames(msgenepop_i22) = c("barcode_id", "BT10", "B96", "BTMS0059", "BTMS0081",
                        "BL13", "BTMS0062", "B126", "BTERN01","B124", "BTMS0057", 
                        "BT30", "B10", "BTMS0083", "BTMS0073", "BT28")


#select and join alleles

#sort allele table by barcode ID
msgenepop_i22 <- msgenepop_i22[order(msgenepop_i22$barcode_id),]
msgenepop_i22$site = substring(msgenepop_i22$barcode_id, 1, 1)

haplotypes_i22 <- as.data.frame(paste(msgenepop_i22$site, ","," ", 
                                      msgenepop_i22$BT10," ", msgenepop_i22$B96, " ",
                                      msgenepop_i22$BTMS0059," ", msgenepop_i22$BTMS0081, " ",
                                      msgenepop_i22$BL13," ", msgenepop_i22$BTMS0062, " ",
                                      msgenepop_i22$B126," ", msgenepop_i22$BTERN01, " ",
                                      msgenepop_i22$B124," ", msgenepop_i22$BTMS0057," ",
                                      msgenepop_i22$BT30," ", msgenepop_i22$B10, " ",
                                      msgenepop_i22$BTMS0083," ", msgenepop_i22$BTMS0073," ",
                                      msgenepop_i22$BT28,
                                  sep = ""))

#Build the Genepop format
sink("data/genepop/ms_genepop_format_impatiens2022.txt")
cat("Title: B. impatiens workers (2022) \n")
cat("BT10 \n")
cat("B96 \n")
cat("BTMS0059 \n")
cat("BTMS0081 \n")
cat("BL13 \n")
cat("BTMS0062 \n")
cat("B126 \n")
cat("BTERN01 \n")
cat("B124 \n")
cat("BTMS0057 \n")
cat("BT30 \n")
cat("B10 \n")
cat("BTMS0083 \n")
cat("BTMS0073 \n")
cat("BT28 \n")
cat("Pop \n")
invisible(apply(as.data.frame(haplotypes_i22[substring(haplotypes_i22[,1],1,1) == "E",]), 1,function(x) cat(x,"\n")))
cat("Pop \n")
invisible(apply(as.data.frame(haplotypes_i22[substring(haplotypes_i22[,1],1,1) == "W",]), 1,function(x) cat(x,"\n")))
cat("Pop \n")
invisible(apply(as.data.frame(haplotypes_i22[substring(haplotypes_i22[,1],1,1) == "S",]), 1,function(x) cat(x,"\n")))
cat("Pop \n")
invisible(apply(as.data.frame(haplotypes_i22[substring(haplotypes_i22[,1],1,1) == "N",]), 1,function(x) cat(x,"\n")))
cat("Pop \n")
invisible(apply(as.data.frame(haplotypes_i22[substring(haplotypes_i22[,1],1,1) == "P",]), 1,function(x) cat(x,"\n")))
cat("Pop \n")
invisible(apply(as.data.frame(haplotypes_i22[substring(haplotypes_i22[,1],1,1) == "H",]), 1,function(x) cat(x,"\n")))
cat("Pop \n")
sink()


# then again, for 2023!
#combine columns for each locus
ms_i23 = data.table(itestalleles2023)
ms_i23[ms_i23 ==0] <- "000"


i1 <- seq(1, length(ms_i23)-1, 2)
i2 <- seq(2, length(ms_i23)-1, 2)
msgenepop_i23 = ms_i23[, Map(paste,
                             .SD[, i1, with = FALSE], .SD[, i2, with = FALSE], 
                             MoreArgs = list(sep="")), 
                       by = "barcode_id"]
colnames(msgenepop_i23) = c("barcode_id", "BT10", "B96", "BTMS0059", "BTMS0081",
                            "BL13", "BTMS0062", "B126", "BTERN01","B124", "BTMS0057", 
                            "BT30", "B10", "BTMS0083", "BTMS0073", "BT28")


#select and join alleles

#sort allele table by barcode ID
msgenepop_i23 <- msgenepop_i23[order(msgenepop_i23$barcode_id),]
msgenepop_i23$site = substring(msgenepop_i23$barcode_id, 1, 1)

haplotypes_i23 <- as.data.frame(paste(msgenepop_i23$site, ","," ", 
                                      msgenepop_i23$BT10," ", msgenepop_i23$B96, " ",
                                      msgenepop_i23$BTMS0059," ", msgenepop_i23$BTMS0081, " ",
                                      msgenepop_i23$BL13," ", msgenepop_i23$BTMS0062, " ",
                                      msgenepop_i23$B126," ", msgenepop_i23$BTERN01, " ",
                                      msgenepop_i23$B124," ", msgenepop_i23$BTMS0057," ",
                                      msgenepop_i23$BT30," ", msgenepop_i23$B10, " ",
                                      msgenepop_i23$BTMS0083," ", msgenepop_i23$BTMS0073," ",
                                      msgenepop_i23$BT28,
                                      sep = ""))

#Build the Genepop format
sink("data/genepop/ms_genepop_format_impatiens2023.txt")
cat("Title: B. impatiens workers (2023) \n")
cat("BT10 \n")
cat("B96 \n")
cat("BTMS0059 \n")
cat("BTMS0081 \n")
cat("BL13 \n")
cat("BTMS0062 \n")
cat("B126 \n")
cat("BTERN01 \n")
cat("B124 \n")
cat("BTMS0057 \n")
cat("BT30 \n")
cat("B10 \n")
cat("BTMS0083 \n")
cat("BTMS0073 \n")
cat("BT28 \n")
cat("Pop \n")
invisible(apply(as.data.frame(haplotypes_i23[substring(haplotypes_i23[,1],1,1) == "E",]), 1,function(x) cat(x,"\n")))
cat("Pop \n")
invisible(apply(as.data.frame(haplotypes_i23[substring(haplotypes_i23[,1],1,1) == "W",]), 1,function(x) cat(x,"\n")))
cat("Pop \n")
invisible(apply(as.data.frame(haplotypes_i23[substring(haplotypes_i23[,1],1,1) == "S",]), 1,function(x) cat(x,"\n")))
cat("Pop \n")
invisible(apply(as.data.frame(haplotypes_i23[substring(haplotypes_i23[,1],1,1) == "N",]), 1,function(x) cat(x,"\n")))
cat("Pop \n")
invisible(apply(as.data.frame(haplotypes_i23[substring(haplotypes_i23[,1],1,1) == "P",]), 1,function(x) cat(x,"\n")))
cat("Pop \n")
invisible(apply(as.data.frame(haplotypes_i23[substring(haplotypes_i23[,1],1,1) == "H",]), 1,function(x) cat(x,"\n")))
cat("Pop \n")
sink()

######################################
# Run genepop on samples
#####################################

#test for HWE
test_HW("data/genepop/ms_genepop_format_impatiens2022.txt", which="Proba", "data/genepop/impatiensHWE2022.txt.D")
test_HW("data/genepop/ms_genepop_format_impatiens2023.txt", which="Proba", "data/genepop/impatiensHWE2023.txt.D")

#test for LD
test_LD("data/genepop/ms_genepop_format_impatiens2022.txt","data/genepop/impatiensLD2022.txt.DIS")
test_LD("data/genepop/ms_genepop_format_impatiens2023.txt","data/genepop/impatiensLD2023.txt.DIS")

```

```{r read in Fis values for each species, echo = FALSE, include = FALSE, cache = TRUE}

# first, mixtus 2022
mix2022HWElines = readLines("data/genepop/mixtusHWE2022.txt.D")
split_mixHWE2022 <- strsplit(mix2022HWElines, "\\s+")

# pull out the estimates for each site
mixHWE2022_east = as.data.frame(do.call(rbind, split_mixHWE2022[269:281]))
mixHWE2022_east$site = "ED"
mixHWE2022_west = as.data.frame(do.call(rbind, split_mixHWE2022[294:306]))
mixHWE2022_west$site = "W"
mixHWE2022_south = as.data.frame(do.call(rbind, split_mixHWE2022[319:331]))
mixHWE2022_south$site = "SD"
mixHWE2022_harvie = as.data.frame(do.call(rbind, split_mixHWE2022[394:406]))
mixHWE2022_harvie$site = "HR"
mixHWE2022_nico = as.data.frame(do.call(rbind, split_mixHWE2022[344:356]))
mixHWE2022_nico$site = "NR"
mixHWE2022_pitt = as.data.frame(do.call(rbind, split_mixHWE2022[369:381]))
mixHWE2022_pitt$site = "PM"

mixHWE2022 = rbind(mixHWE2022_east, mixHWE2022_harvie, mixHWE2022_nico, mixHWE2022_pitt, mixHWE2022_south, mixHWE2022_west)
colnames(mixHWE2022) = c("locus", "pval", "SE", "FisWC", "FisRH", "steps", "switches", "site")
mixHWE2022$year = 2022



## next, for mixtus 2023
mix2023HWElines = readLines("data/genepop/mixtusHWE2023.txt.D")
split_mixHWE2023 <- strsplit(mix2023HWElines, "\\s+")

# pull out the estimates for each site
mixHWE2023_east = as.data.frame(do.call(rbind, split_mixHWE2023[269:281]))
mixHWE2023_east$site = "ED"
mixHWE2023_west = as.data.frame(do.call(rbind, split_mixHWE2023[294:306]))
mixHWE2023_west$site = "W"
mixHWE2023_south = as.data.frame(do.call(rbind, split_mixHWE2023[319:331]))
mixHWE2023_south$site = "SD"
mixHWE2023_harvie = as.data.frame(do.call(rbind, split_mixHWE2023[394:406]))
mixHWE2023_harvie$site = "HR"
mixHWE2023_nico = as.data.frame(do.call(rbind, split_mixHWE2023[344:356]))
mixHWE2023_nico$site = "NR"
mixHWE2023_pitt = as.data.frame(do.call(rbind, split_mixHWE2023[369:381]))
mixHWE2023_pitt$site = "PM"

mixHWE2023 = rbind(mixHWE2023_east, mixHWE2023_harvie, mixHWE2023_nico, mixHWE2023_pitt, mixHWE2023_south, mixHWE2023_west)
colnames(mixHWE2023) = c("locus", "pval", "SE", "FisWC", "FisRH", "steps", "switches", "site")
mixHWE2023$year = 2023

mixHWE = rbind(mixHWE2022, mixHWE2023)


## next, for impatiens 2022!
imp2022HWElines = readLines("data/genepop/impatiensHWE2022.txt.D")
split_impHWE2022 <- strsplit(imp2022HWElines, "\\s+")

# pull out the estimates for each site
impHWE2022_east = as.data.frame(do.call(rbind, split_impHWE2022[305:319]))
impHWE2022_east$site = "ED"
impHWE2022_west = as.data.frame(do.call(rbind, split_impHWE2022[332:346]))
impHWE2022_west$site = "W"
impHWE2022_south = as.data.frame(do.call(rbind, split_impHWE2022[359:373]))
impHWE2022_south$site = "SD"
impHWE2022_harvie = as.data.frame(do.call(rbind, split_impHWE2022[440:454]))
impHWE2022_harvie$site = "HR"
impHWE2022_nico = as.data.frame(do.call(rbind, split_impHWE2022[386:400])[,1:7])
impHWE2022_nico$site = "NR"
impHWE2022_pitt = as.data.frame(do.call(rbind, split_impHWE2022[413:427]))
impHWE2022_pitt$site = "PM"

impHWE2022 = rbind(impHWE2022_east, impHWE2022_harvie, impHWE2022_nico, impHWE2022_pitt, impHWE2022_south, impHWE2022_west)
colnames(impHWE2022) = c("locus", "pval", "SE", "FisWC", "FisRH", "steps", "switches", "site")
impHWE2022$year = 2022



## repeat for impatiens 2023!
imp2023HWElines = readLines("data/genepop/impatiensHWE2023.txt.D")
split_impHWE2023 <- strsplit(imp2023HWElines, "\\s+")

# pull out the estimates for each site
impHWE2023_east = as.data.frame(do.call(rbind, split_impHWE2023[305:319]))
impHWE2023_east$site = "ED"
impHWE2023_west = as.data.frame(do.call(rbind, split_impHWE2023[332:346]))
impHWE2023_west$site = "W"
impHWE2023_south = as.data.frame(do.call(rbind, split_impHWE2023[359:373]))
impHWE2023_south$site = "SD"
impHWE2023_harvie = as.data.frame(do.call(rbind, split_impHWE2023[440:454]))
impHWE2023_harvie$site = "HR"
impHWE2023_nico = as.data.frame(do.call(rbind, split_impHWE2023[386:400]))
impHWE2023_nico$site = "NR"
impHWE2023_pitt = as.data.frame(do.call(rbind, split_impHWE2023[413:427]))
impHWE2023_pitt$site = "PM"

impHWE2023 = rbind(impHWE2023_east, impHWE2023_harvie, impHWE2023_nico, impHWE2023_pitt, impHWE2023_south, impHWE2023_west)
colnames(impHWE2023) = c("locus", "pval", "SE", "FisWC", "FisRH", "steps", "switches", "site")
impHWE2023$year = 2023

impHWE = rbind(impHWE2022, impHWE2023)



##############################################
# Make sure all columns are right data format
##############################################

impHWE$FisWC = as.numeric(impHWE$FisWC)
impHWE$locus = factor(impHWE$locus)
impHWE$site = factor(impHWE$site)

mixHWE$FisWC = as.numeric(mixHWE$FisWC)
mixHWE$locus = factor(mixHWE$locus)
mixHWE$site = factor(mixHWE$site)

```

```{r plot Fis distributions for each species, fig.width = 6, fig.height = 3.25, fig.align = "center", echo = FALSE, cache = TRUE}
# combine data frames
impHWE$species = "B. impatiens"
mixHWE$species = "B. mixtus"
HWE = rbind(impHWE, mixHWE)

ggplot(HWE, aes(x = FisWC)) +
  geom_histogram(binwidth = 0.02) +
  geom_vline(xintercept = 0, color = "blue", linetype = "dotted") +
  geom_vline(xintercept = -0.1, color = "red", linetype = "dotted") +
  geom_vline(xintercept = 0.1, color = "red", linetype = "dotted") +
  theme_minimal() +
  labs(y = "Count", x = expression(F[is] * " (W&C)"), title = expression(F[is] * " Distributions Across Sites and Species")) +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(axis.text.x = element_text(size = 4)) +
  facet_grid(species ~ site)

# ggplot(HWE, aes(x = FisWC)) +
#   geom_histogram() +
#   geom_vline(xintercept = 0, color = "red", linetype = "dotted") +
#   theme_minimal() +
#   facet_grid(species ~ 1)


```

We fit linear models to $F_{is}$ estimates with locus and site as fixed predictors. We used sum-to-zero coding so that model intercepts represented mean $F_{is}$ across all loci, and calculated the estimated marginal mean of each locus using the *emmeans* package [@lenthEmmeansEstimatedMarginal2024]. Marginal means are plotted below; the dashed black line denotes $F_{is} = 0$, and the dashed red line denotes mean $F_{is}$ for each species.

```{r initial linear models of Fis, fig.width = 6, fig.height = 2.5, fig.align = "center", echo=FALSE, cache = TRUE}
#############################################
# Initial models and plots for impatiens
#############################################
# set site and locus to sum-to-zero contrasts
contrasts(impHWE$locus) = contr.sum
contrasts(impHWE$site) = contr.sum

# fit the model
model <- lm(FisWC ~ locus + site, data = impHWE)

# get locus vs grand mean comparisons
emm_locus = emmeans(model, "locus")
comp_locus = contrast(emm_locus, method = "eff", adjust = "none")
#summary(comp_locus)
#summary(model)

# plot locus effects
emm_df <- as.data.frame(emm_locus)
impatiens_locus_init = ggplot(emm_df, aes(x = locus, y = emmean)) +
  geom_point() +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_hline(yintercept = model$coefficients[1], linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(y = "", x = "Locus", title = substitute(italic("B. impatiens"))) +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(axis.text.x = element_text(size = 8, angle = 90, hjust = 1))

#############################################
# Initial models and plots for mixtus
#############################################
# set site and locus to sum-to-zero contrasts
contrasts(mixHWE$locus) = contr.sum
contrasts(mixHWE$site) = contr.sum

# fit the model
model <- lm(FisWC ~ locus + site, data = mixHWE)

# get locus vs grand mean comparisons
emm_locus = emmeans(model, "locus")
comp_locus = contrast(emm_locus, method = "eff", adjust = "none")
#summary(comp_locus)
#summary(model)

# plot locus effects
emm_df <- as.data.frame(emm_locus)
mixtus_locus_init = ggplot(emm_df, aes(x = locus, y = emmean)) +
  geom_point() +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_hline(yintercept = model$coefficients[1], linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(y = expression("Estimated mean " * F[is] * " (W&C)"), x = "Locus", title = substitute(italic("B. mixtus"))) +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(axis.text.x = element_text(size = 8, angle = 90, hjust = 1))

grid.arrange(mixtus_locus_init, impatiens_locus_init, ncol = 2)
```

We iteratively removed loci with $F_{is}$ significantly different from the species mean $F_{is}$, starting with the locus with the greatest deviation, and re-running the model after each removal (i.e., because removing a locus with a high or low inbreeding coefficient will change the species mean estimate and therefore all comparisons to the mean). We did not apply an adjustment for multiple-hypothesis testing, but instead utilized a relatively stringent p-value ($\alpha$ = 0.01) for removal of loci.

```{r, iteratively remove poor loci for impatiens, echo = FALSE, cache = TRUE}
#############################################
# Remove poor loci for impatiens
#############################################
##### remove BTMS0073
impHWE_trunc = impHWE[impHWE$locus != "BTMS0073",]

# reset factor levels
impHWE_trunc$locus = factor(impHWE_trunc$locus)

# set site and locus to sum-to-zero contrasts
contrasts(impHWE_trunc$locus) = contr.sum
contrasts(impHWE_trunc$site) = contr.sum

# fit the model
model <- lm(FisWC ~ locus + site, data = impHWE_trunc)

# get locus vs grand mean comparisons
emm_locus = emmeans(model, "locus")
comp_locus = contrast(emm_locus, method = "eff", adjust = "none")


##### remove BT28
impHWE_trunc = impHWE_trunc[impHWE_trunc$locus != "BT28",]

# reset factor levels
impHWE_trunc$locus = factor(impHWE_trunc$locus)

# set site and locus to sum-to-zero contrasts
contrasts(impHWE_trunc$locus) = contr.sum
contrasts(impHWE_trunc$site) = contr.sum

# fit the model
model <- lm(FisWC ~ locus + site, data = impHWE_trunc)

# get locus vs grand mean comparisons
emm_locus = emmeans(model, "locus")
comp_locus = contrast(emm_locus, method = "eff", adjust = "none")
#summary(model)
#summary(comp_locus)

# plot locus effects
emm_df <- as.data.frame(emm_locus)
impatiens_locus = ggplot(emm_df, aes(x = locus, y = emmean)) +
  geom_point() +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_hline(yintercept = model$coefficients[1], linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(y = "", x = "Locus", title = substitute(italic("B. impatiens"))) +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(axis.text.x = element_text(size = 8, angle = 90, hjust = 1))


```


```{r, iteratively remove poor loci for mixtus, echo = FALSE, cache = TRUE}
#############################################
# Remove poor loci for mixtus
#############################################
##### remove BTERN01
mixHWE_trunc = mixHWE[mixHWE$locus != "BTERN01",]

# reset factor levels
mixHWE_trunc$locus = factor(mixHWE_trunc$locus)

# set site and locus to sum-to-zero contrasts
contrasts(mixHWE_trunc$locus) = contr.sum
contrasts(mixHWE_trunc$site) = contr.sum

# fit the model
model <- lm(FisWC ~ locus + site, data = mixHWE_trunc)

# get locus vs grand mean comparisons
emm_locus = emmeans(model, "locus")
comp_locus = contrast(emm_locus, method = "eff", adjust = "none")


##### remove BTMS0104
mixHWE_trunc = mixHWE_trunc[mixHWE_trunc$locus != "BTMS0104",]

# reset factor levels
mixHWE_trunc$locus = factor(mixHWE_trunc$locus)

# set site and locus to sum-to-zero contrasts
contrasts(mixHWE_trunc$locus) = contr.sum
contrasts(mixHWE_trunc$site) = contr.sum

# fit the model
model <- lm(FisWC ~ locus + site, data = mixHWE_trunc)

# get locus vs grand mean comparisons
emm_locus = emmeans(model, "locus")
comp_locus = contrast(emm_locus, method = "eff", adjust = "none")


##### remove BTMS0059
mixHWE_trunc = mixHWE_trunc[mixHWE_trunc$locus != "BTMS0059",]

# reset factor levels
mixHWE_trunc$locus = factor(mixHWE_trunc$locus)

# set site and locus to sum-to-zero contrasts
contrasts(mixHWE_trunc$locus) = contr.sum
contrasts(mixHWE_trunc$site) = contr.sum

# fit the model
model <- lm(FisWC ~ locus + site, data = mixHWE_trunc)

# get locus vs grand mean comparisons
emm_locus = emmeans(model, "locus")
comp_locus = contrast(emm_locus, method = "eff", adjust = "none")
#summary(model)
#summary(comp_locus)

# plot locus effects
emm_df <- as.data.frame(emm_locus)
mixtus_locus = ggplot(emm_df, aes(x = locus, y = emmean)) +
  geom_point() +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_hline(yintercept = model$coefficients[1], linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(y = expression("Estimated mean " * F[is] * " (W&C)"), x = "Locus", title = substitute(italic("B. mixtus"))) +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(axis.text.x = element_text(size = 8, angle = 90, hjust = 1))

```

This process resulted in the removal of loci BTERN01, BTMS0104, and BTMS0059 from downstream analyses for *B. mixtus* and removal of BTMS0073 and BT28 for *B. impatiens*. $F_{is}$ estimates for the remaining loci (n = 10 for *B. mixtus*, n = 13 for *B. impatiens*) are shown below.

```{r plots of remaining loci after Fis removals, fig.width = 6, fig.height = 2.5, fig.align = "center", echo = FALSE, cache = TRUE}
grid.arrange(mixtus_locus, impatiens_locus, ncol =2)

```
## Check markers for linkage disequilibrium
Next, we checked locus pairs for linkage disequilibrium (LD). Marker linkage can lead to non-independent assortment, a condition which violates the assumptions of COLONY. LD was calculated for each locus pair in each population (site, year, species groups) in *genepop*. We applied a Bonferroni correction for multiple hypothesis testing, and flagged locus pairs which showed significant deviations in $>2$ populations.

[[Is there a better way of approaching this? It seems odd to me to correct for multiple hypothesis testing and then only drop loci if they exceed the Bonferroni-correct p-value in *multiple* cases--it's common practice in the literature, but perhaps it would be better to first assess $F_{st}$ between sites and years and then test LD of loci across *all* samples rather than individually for each population?]]

```{r check for LD in genepop, echo = FALSE, cache = TRUE}

##############################
# IMPATIENS
#############################

LD2022lines = readLines("data/genepop/impatiensLD2022.txt.DIS")
split_linesLD2022 <- strsplit(LD2022lines, "\\s+")
LD2022 = do.call(rbind, split_linesLD2022[13:644])
colnames(LD2022) = LD2022[1,]
colnames(LD2022)[colnames(LD2022) == 'P-Value'] = "pval"
LD2022 = as.data.frame(LD2022[3:nrow(LD2022),])
LD2022$year = 2022

LD2023lines = readLines("data/genepop/impatiensLD2023.txt.DIS")
split_linesLD2023 <- strsplit(LD2023lines, "\\s+")
LD2023 = do.call(rbind, split_linesLD2023[13:644])
colnames(LD2023) = LD2023[1,]
colnames(LD2023)[colnames(LD2023) == 'P-Value'] = "pval"
LD2023 = as.data.frame(LD2023[3:nrow(LD2023),])
LD2023$year = 2023

impatiensLD = rbind(LD2022, LD2023)

# check out loci that are below Bonferroni-corrected P-value
alpha_imp = 0.05/nrow(impatiensLD)

impatiensLD = impatiensLD[as.numeric(impatiensLD$pval) < alpha_imp,]
impatiensLD = impatiensLD[!(impatiensLD$`Locus#1`%in% c("BTMS0073", "BT28")),]
impatiensLD = impatiensLD[!(impatiensLD$`Locus#2`%in% c("BTMS0073", "BT28")),]


##############################
# MIXTUS
#############################

LD2022lines = readLines("data/genepop/mixtusLD2022.txt.DIS")
split_linesLD2022 <- strsplit(LD2022lines, "\\s+")
LD2022 = do.call(rbind, split_linesLD2022[13:482])
colnames(LD2022) = LD2022[1,]
colnames(LD2022)[colnames(LD2022) == 'P-Value'] = "pval"
LD2022 = as.data.frame(LD2022[3:nrow(LD2022),])
LD2022$year = 2022

LD2023lines = readLines("data/genepop/mixtusLD2023.txt.DIS")
split_linesLD2023 <- strsplit(LD2023lines, "\\s+")
LD2023 = do.call(rbind, split_linesLD2023[13:482])
colnames(LD2023) = LD2023[1,]
colnames(LD2023)[colnames(LD2023) == 'P-Value'] = "pval"
LD2023 = as.data.frame(LD2023[3:nrow(LD2023),])
LD2023$year = 2023

mixtusLD = rbind(LD2022, LD2023)

# check out loci that are below Bonferroni-corrected P-value
alpha_mix = 0.05/nrow(mixtusLD)

mixtusLD = mixtusLD[as.numeric(mixtusLD$pval) < alpha_mix,]
mixtusLD = mixtusLD[!(mixtusLD$`Locus#1`%in% c("BTMS0059", "BTMS0104", "BTERN01")),]
mixtusLD = mixtusLD[!(mixtusLD$`Locus#2`%in% c("BTMS0059", "BTMS0104", "BTERN01")),]


```

There was not strong support for linkage disequilibrium between locus pairs in *B. mixtus.* Four pairs showed signs of LD, but each occurred at only a single site in a single year. For *B. impatiens*, there was strong evidence for linkage disequilibrium between BTMS0057 and BL13 (7 out of 12 populations showed significant LD following Bonferroni correction). To determine which locus to maintain, we calculated the polymorphic information content (PIC) using the package *PopGenUtils* [@tourvasPopGenUtilsCollectionUseful2025].


```{r calculate PIC of BTMS0057 and BL13 in impatiens, include = FALSE, cache = TRUE}

# load genepop files
file.copy("data/genepop/ms_genepop_format_impatiens2022.txt",
          "data/genepop/ms_genepop_format_impatiens2022.gen")
file.copy("data/genepop/ms_genepop_format_impatiens2023.txt",
          "data/genepop/ms_genepop_format_impatiens2023.gen")

genind_imp2022 <- read.genepop("data/genepop/ms_genepop_format_impatiens2022.gen", ncode = 3)
genind_imp2023 <- read.genepop("data/genepop/ms_genepop_format_impatiens2023.gen", ncode = 3)

# calculate PIC
pic_values_2022 <- pic_calc(genind_imp2022, ploidy = 2, bypop = FALSE)
pic_values_2023 <- pic_calc(genind_imp2023, ploidy = 2, bypop = FALSE)

#pic_values_2022
#pic_values_2023

# repeat for mixtus
# load genepop files
file.copy("data/genepop/ms_genepop_format_mixtus2022.txt",
          "data/genepop/ms_genepop_format_mixtus2022.gen")
file.copy("data/genepop/ms_genepop_format_mixtus2023.txt",
          "data/genepop/ms_genepop_format_mixtus2023.gen")

genind_mix2022 <- read.genepop("data/genepop/ms_genepop_format_mixtus2022.gen", ncode = 3)
genind_mix2023 <- read.genepop("data/genepop/ms_genepop_format_mixtus2023.gen", ncode = 3)

# calculate PIC
pic_values_2022 <- pic_calc(genind_mix2022, ploidy = 2, bypop = FALSE)
pic_values_2023 <- pic_calc(genind_mix2023, ploidy = 2, bypop = FALSE)

#pic_values_2022
#pic_values_2023

```
We found that BTMS0057 had higher PIC in both years (PIC = 0.78) compared to BL13 (PIC = 0.47 in 2022 and 0.45 in 2023). For this reason, we chose to maintain BTMS0057 and remove BL13 from further analyses for *B. impatiens*.


## Calculate $F_{st}$ and $F_{is}$ between sites
After locus quality screening, we calculated global and pairwise $F_{st}$ following @nei1987molecular in the package *hierfstat* [@goudetHierfstatPackageCompute2005].

```{r calculate Fst between sites, include = FALSE, cache = TRUE}
# load mixtus genepop files and make genind!
file.copy("data/genepop/ms_genepop_format_mixtus2022.txt",
          "data/genepop/ms_genepop_format_mixtus2022.gen")
file.copy("data/genepop/ms_genepop_format_mixtus2023.txt",
          "data/genepop/ms_genepop_format_mixtus2023.gen")

genind_mix2022 <- read.genepop("data/genepop/ms_genepop_format_mixtus2022.gen", ncode = 3)
genind_mix2023 <- read.genepop("data/genepop/ms_genepop_format_mixtus2023.gen", ncode = 3)

# convert all to hierfstat objects
hf_imp2022 <- genind2hierfstat(genind_imp2022)
hf_imp2023 <- genind2hierfstat(genind_imp2023)
hf_mix2022 <- genind2hierfstat(genind_mix2022)
hf_mix2023 <- genind2hierfstat(genind_mix2023)

# remove dropped loci!
hf_imp2022 = hf_imp2022[,!colnames(hf_imp2022) %in% c("BT28", "BTMS0073", "BL13")]
hf_imp2023 = hf_imp2023[,!colnames(hf_imp2023) %in% c("BT28", "BTMS0073", "BL13")]
hf_mix2022 = hf_mix2022[,!colnames(hf_mix2022) %in% c("BTMS0104", "BTMS0059", "BTERN01")]
hf_mix2023 = hf_mix2023[,!colnames(hf_mix2023) %in% c("BTMS0104", "BTMS0059", "BTERN01")]

# set individuals with partially missing data (e.g., 1 called peak, 1 unbinned peak) to NA
handle_partial = function(hf_obj){
  for (col in 2:ncol(hf_obj)){
    for (row in 1:nrow(hf_obj)){
      if (is.na(hf_obj[row, col])){
        # skip
        } else if (nchar(hf_obj[row, col]) < 6){
        hf_obj[row, col] = NA
      }
    }
  }
  return(hf_obj)
}

hf_imp2022 = handle_partial(hf_imp2022)
hf_imp2023 = handle_partial(hf_imp2023)
hf_mix2022 = handle_partial(hf_mix2022)
hf_mix2023 = handle_partial(hf_mix2023)


# calculate global Fst
fst_imp2022 <- basic.stats(hf_imp2022)$overall["Fst"]
fst_imp2023 <- basic.stats(hf_imp2023)$overall["Fst"]
fst_mix2022 <- basic.stats(hf_mix2022)$overall["Fst"]
fst_mix2023 <- basic.stats(hf_mix2023)$overall["Fst"]

# make table
fst = data.frame(Species = c("B. mixtus", "B. mixtus", "B. impatiens", "B.impatiens"),
                    Year = c(2022, 2023, 2022, 2023),
                    Fst = c(fst_mix2022, fst_mix2023, fst_imp2022, fst_imp2023))
fst

```


```{r calculate pairwise Fst, include = FALSE, cache = TRUE}
pfst_imp2022 <- pairwise.neifst(hf_imp2022)
pfst_imp2023 <- pairwise.neifst(hf_imp2023)
pfst_mix2022 <- pairwise.neifst(hf_mix2022)
pfst_mix2023 <- pairwise.neifst(hf_mix2023)

max(pfst_imp2022, na.rm = TRUE)
max(pfst_imp2023, na.rm = TRUE)
max(pfst_mix2022, na.rm = TRUE)
max(pfst_mix2023, na.rm = TRUE)
```

For both species and years, estimates of global $F_{st}$ were less than 0.005. Pairwise $F_{st}$ ranged from -0.002** to 0.01, indicating little or no genetic differentiation between surveyed sub-populations. [[**I read that $F_{st}$ can sometimes be negative if between population variance is estimated to be slightly smaller than within population variance, as an artifact of finite sampling--would you agree, and if so is it better to report like this or round to 0?]]

Finally, we computed the marginal mean $F_{is}$ for each site to determine whether there was evidence for inbreeding following locus removal, and whether it varied between sites.

```{r check for site-level inbreeding, fig.width = 6, fig.height = 2.5, fig.align = "center", echo = FALSE, cache = TRUE}
######################
# IMPATIENS
######################
# fit the model
model <- lm(FisWC ~ site, data = impHWE_trunc)

# get site vs grand mean comparisons
emm_site = emmeans(model, "site")
comp_site = contrast(emm_site, method = "eff", adjust = "none")

# plot site effects
emm_df <- as.data.frame(emm_site)
impatiens_site = ggplot(emm_df, aes(x = site, y = emmean)) +
  geom_point() +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_hline(yintercept = model$coefficients[1], linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(y = expression("Estimated mean " * F[is] * " (W&C)"), x = "Site", title = substitute(italic("B. impatiens"))) +
  theme(axis.text.x = element_text(size = 10))



######################
# MIXTUS
######################
# fit the model
model <- lm(FisWC ~ site, data = mixHWE_trunc)

# get site vs grand mean comparisons
emm_site = emmeans(model, "site")
comp_site = contrast(emm_site, method = "eff", adjust = "none")

# plot site effects
emm_df <- as.data.frame(emm_site)
mixtus_site = ggplot(emm_df, aes(x = site, y = emmean)) +
  geom_point() +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_hline(yintercept = model$coefficients[1], linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(y = expression("Estimated mean " * F[is] * " (W&C)"), x = "Site", title = substitute(italic("B. mixtus"))) +
  theme(axis.text.x = element_text(size = 10))

# plot those puppies
grid.arrange(mixtus_site, impatiens_site, ncol =2)
```

## Finalize siblingship assignments
We used our updated marker sets to more stringently assign siblingships. @jonesCOLONYProgramParentage2010 state that the inbreeding model in COLONY 2.0 should only be implemented for dioecious species when there is strong evidence for high levels of inbreeding. Because we found evidence for only minor inbreeding (e.g., $F_{is} < 0.05$) we ran all final colony assignments using the no-inbreeding model. Given the very low estimates of global and pairwise $F_{st}$, we chose to combine sites for siblingship assignments to maximize the accuracy of allele frequency estimation and to validate our methodology (described below) for excluding spurious siblingships.

[[I have not run the finalized colony assignments yet, bu the method I have come up with for excluding the spurious siblingships is to run COLONY 5 times for each population and only maintain the siblingships which occur with high probability in all of them. I will need to develop a simple heuristic for dealing with non-circular families (e.g., A related to B, B related to C, A not related to C). Currently I am thinking: (1) check if A is related to C in at least 3/5 runs of COLONY, (2) if yes to (1), keep the whole family. If no to (1), randomly draw a sib pair to maintain (AB or BC). This seems to happen only rarely from what I can see, so hopefully it won't be hugely problematic.]]




\newpage


